Evaluation of +S2 sea ice forecast and some other better title

### 2025-02-19 analysis/paper/access-ice.qmd
Integrates some changes into figures

```{r hindcast_extent}
hindcast_extent <- rbind(S1 = readRDS(here::here("data/derived/S1_hindcast_extent.Rds")),
                         S2 = readRDS(here::here("data/derived/S2_hindcast_extent.Rds")), 
                         idcol = "model") |>
  _[, model := relevel(factor(model), "S2")] |> 
  _[, lag := as.numeric(as.Date(time) - forecast_time)] |> 
  _[lag > 0] |> 
  _[!(month(forecast_time) == 1 & member == 7 & model == "S1")]  # This member is wrong
```

```{r}
hindcast_area <- rbind(S1 = readRDS(here::here("data/derived/S1_hindcast_area.Rds")),
                       S2 = readRDS(here::here("data/derived/S2_hindcast_area.Rds")), 
                       idcol = "model") |>
  _[, model := relevel(factor(model), "S2")] |> 
  _[, lag := as.numeric(as.Date(time) - forecast_time)] |> 
  _[lag > 0] |> 
  _[!(month(forecast_time) == 1 & member == 7 & model == "S1")]  # This member is wrong
```

@@

months <- 1:12
hindcast_clim <- lapply(setNames(months, months), \(month) {
  here::here(glue::glue("data/derived/hindcast_climatology/mo_aice_1981{zeropad(month)}01_mean.nc")) |> 
    ReadNetCDF("aice")
}) |> 
  rbindlist(idcol = "forecast_month") |> 
  _[, forecast_month := as.numeric(forecast_month)] |> 
  _[, forecast_time := lubridate::make_datetime(1981, forecast_month, 01)] |> 
  _[, time := lubridate::floor_date(time, "month")] |> 
  _[, lag := mondf(forecast_time, time)] |> 
  _[lag == 0] |> 
  _[, time2 := update(time, year = 2000)]

obs_clim <- CDR() |> 
  climatology() |> 
  cdo_ymonmean() |>
  cdo_execute() |> 
  ReadNetCDF(c(obs = "aice")) |> 
  setnames("time", "time2") |> 
  _[, time2 := update(time2, day = 1)]

hindcast_clim <- hindcast_clim |> 
  merge(obs_clim) |> 
  _[, month := month(time)] 
```

@@

hindcast_clim |> 
  # dcast(time + xgrid + ygrid ~ model, value.var = "aice") |> 
  # setnames(which_dataset, "obs") |>
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = aice - obs, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised(paste0("Bias with respect to ",
                                          labels_models[which_dataset]),
                                   low = scales::muted("red"),
                                   high = scales::muted("blue")) +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ lubridate::month(time, label = TRUE)) +
  wide_legend +
  tag_facets()
```

@@

monthly_extent <- extent_daily |> 
  _[aice == 0, aice := NA]  |> 
  _[, .(aice = mean(aice)), by = .(dataset, time = as.Date(round_date(time, "month")))] |>
  _[, aice := Anomaly(aice, year(time) %between% c(1981, 2011), na.rm = TRUE), 
    by = .(month(time), dataset)]

@@

hindcast_extent |> 
  copy() |> 
  # _[model == "S2"] |> 
  _[, aice := aice - mean(aice[year(time) %between% c(1981, 2011)]), 
    by = .(lag, model, time = update(time, year = 2000))] |> 
  _[, .(aice = mean(aice)), by = .(model, forecast_time, time = as.Date(round_date(time, "month")))] |> 
  _[, lag := months_difference(time, forecast_time) + 1] |> 
  _[lag %in% c(1, 3, 5, 7)] |> 
  ggplot(aes(time, aice)) +
  geom_line(data = monthly_extent[year(time) <= max(year(hindcast_extent$time))] |> 
              _[dataset == which_dataset],
            aes(group = dataset, color = dataset)) +
  # geom_smooth(data = monthly_extent[year(time) <= max(year(hindcast_extent$time))] |> 
  #               _[dataset == "cdr"], 
  #             aes(group = dataset, color = dataset),
  #             se = FALSE, span = 10*12/N, n = 366) +
  geom_line(aes(color = model))  +
  # geom_smooth(aes(color = model), se = FALSE, span = 10*12/N, n = 366) +
  scale_y_continuous(NULL,
                     labels = labels_extent) +
  scale_x_date(NULL, expand = c(0, 0)) +
  facet_grid(lag ~ model, labeller = labeller(lag = \(x) paste0("Lag: ", x))) +
  scale_color_models
```

>>>>

```{r hindcast_extent}
hindcast_extent <- fread(here::here("data/derived/hindcast_extent.csv")) |> 
  _[, model := relevel(factor(model), "S2")] |> 
  _[, lag := as.numeric(as.Date(time) - as.Date(forecast_time))] |> 
  _[lag > 0] |> 
  _[!(month(forecast_time) == 1 & member == 7 & model == "S1")]  # This member is wrong
```

@@

months <- 1:12

hindcast_clim <- lapply(setNames(months, months), \(month) {
  here::here(glue::glue("data/derived/climatology/S2/{zeropad(month)}/em.nc")) |> 
    cdo_selmonth(month) |> 
    cdo_execute() |> 
    ReadNetCDF("aice")
}) |> 
  rbindlist(idcol = "forecast_month") |> 
  _[, forecast_month := as.numeric(forecast_month)]

hindcast_clim <- hindcast_clim |> 
  copy() |> 
  _[, .(aice = mean(aice)), by = .(month(time), ygrid, xgrid, forecast_month)] |> 
  _[forecast_month == month]

obs_clim <- CDR() |> 
  climatology() |> 
  cdo_ymonmean() |>
  cdo_execute() |> 
  ReadNetCDF(c(obs = "aice")) |> 
  _[, month := month(time)]

hindcast_clim <- hindcast_clim |> 
  merge(obs_clim)
```

@@

hindcast_clim |> 
  # dcast(time + xgrid + ygrid ~ model, value.var = "aice") |> 
  # setnames(which_dataset, "obs") |>
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = aice - obs, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised(paste0("Bias with respect to ",
                                          labels_models[which_dataset]),
                                   low = scales::muted("red"),
                                   high = scales::muted("blue")) +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ month) +
  wide_legend +
  tag_facets()
```

@@

monthly_extent <- extent_daily |> 
  _[dataset == which_dataset] |> 
  _[aice == 0, aice := NA]  |> 
  _[, .(aice = mean(aice)), by = .(dataset, time = as.Date(round_date(time, "month")))] |>
  _[, aice := Anomaly(aice, year(time) %between% c(1981, 2011), na.rm = TRUE), 
    by = .(month(time), dataset)]

@@

s1_range <- hindcast_extent[model == "S1"][, range(time)]
s2_range <- hindcast_extent[model == "S2"][, range(time)]

hindcast_extent |> 
  copy() |> 
  _[, aice := aice - mean(aice[year(time) %between% c(1981, 2011)]), 
    by = .(lag, model, time = update(time, year = 2000))] |> 
  _[, .(forecast = mean(aice)), by = .(model, forecast_time, time = as.Date(round_date(time, "month")))] |> 
  _[, lag := months_difference(time, forecast_time)] |> 
  _[, merge(.SD, monthly_extent, all = TRUE), by = .(model, lag)] |> 
  _[lag %in% (c(1, 3, 5, 7) - 1)] |>
  _[time %between% s2_range] |> 
  ggplot(aes(time)) +
  geom_line(aes(y = aice, group = dataset, color = dataset)) +
  geom_line(aes(y = forecast, color = model))  +
  geom_text(data = \(x) x[time %between% s1_range, correlate(aice, forecast, use = "complete.obs"), by = .(lag, model)],
            aes(label = text), x = -Inf, y = Inf, 
            vjust = 1.2, hjust = -0.2) +
  scale_y_continuous(NULL,
                     labels = labels_extent) +
  scale_x_date(NULL, expand = c(0, 0)) +
  facet_grid(lag ~ model, labeller = labeller(lag = \(x) paste0("Month: ", x))) +
  scale_color_models
```
